Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2024-03-04 paper/markdown/supplementary_material.Rmd
Figure and Table update

Clinically relevant symptoms of anxiety were defined as GAD-7 $\geq$ 10, 
clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 10 [@Humer2022], 
significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008]. 
For a comparison of frequency and severity of depressive symptoms in the study cohort with microcensus data for the general Austrian population, mild depressive symptoms were defined as 5 - 9, moderate symptoms as 10 - 14, moderately severe symptoms as 15 - 19, and severe symptoms as $\geq$ 20 points of the PHQ-8 score [@Kroenke2002a].

@@

Lifetime estimates of prevalence of traumatic events and PTSD in the general population were obtained from representative survey studies [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017]. 
Frequency of traumatic events, manifest PTSD and PTSD symptoms subsumed under the B, C, D, and E domains of the PCL-5 instrument in Swiss mountain rescuers were extracted from the report by Mikutta et al. [@Mikutta2022].

Data on frequency of clinically relevant anxious (GAD-7 $\geq$ 10) and depressive (PHQ-9 $\geq$ 10) symptoms in Austria in 2022 were obtained from a publication by Humer et al. [@Humer2022]. 
Data on frequency and severity of depressive symptoms measured with the PHQ-8 instrument [@Kroenke2002a] in a pre-pandemic nation-wide Austrian 2019 microcensus study were obtained from a publication by Statistik Austria [@StatistikAustria2019a].

Mean scores of resilient comping with their standard deviations (SD) obtained with the RS-13 tool were extracted from reports with samples of the German [@Leppert2008] and Tyrolean [@Chernova2021] populations, and from a cohort of Swiss mountain rescuers [@Mikutta2022].

@@

* r statistic: weak: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5

* Cramer's V statistic: weak: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5

* $\eta^2$ statistic: weak: < 0.06, moderate: 0.06 - 0.14, large: $\geq$ 0.14

* Cohen's $\kappa$ inter-rater reliability statistic: none: < 0.2, minimal: 0.2 - 0.4, weak: 0.4 - 0.6, moderate: 0.6 - 0.8, strong: $\geq$ 0.8

@@

An RData file with anonymized patient data will be made available upon request to the corresponding author. 
The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.

@@

flextable::flextable(suppl_tables$aut_population) %>% 
  set_widths(c(3.5, rep(5, 2), 2.8, 2.8)) %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic, categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           ref_symbols = 'a', 
           part = 'header') %>% 
  theme_vanilla

@@

```{r fig-accident, fig.width = figur::convert(suppl_figures$accident, to = 'in')$w, fig.height = figur::convert(suppl_figures$accident, to = 'in')$h, fig.cap = 'Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.'}

@@

__Supplementary Figure S\@ref(fig:fig-accident). Mountain sport type, accident month, and age of the mountain accident victim in the study cohort and Austria.__

_Mountain sport types (A), accident months (B) and age class distribution (C) were compared between the study cohort and victims of mountain sport accidents in Austria in 2023. Frequencies for the nation-wide data set of mountain accidents were extracted from the annual report of the Austrian Board of Mountain Safety. In the comparison of sport types, non-sportive mountain accidents (e.g. during hunting or forestry) were excluded from the national statistic. Non-classifiable sport types (category 'other') in the study cohort were excluded from the analysis presented in (C). In the comparison of the age class distribution, only individuals older than 20 years were included. Statistical significance of the global differences was determined by $\chi^2$ test with Cramer's V effect size statistic. Pairwise differences for particular sport types, months and age classes were assessed by $\chi^2$ test corrected for multiple testing with the Holm method. Percentages of categories within the complete observation sets are presented as bar plots. Bars are labeled with the corresponding percent values. Total observation numbers are displayed in the plot legends. Global effect sizes and p values are presented in the plot captions. Significant results of post-hoc tests are highlighted with bold X axis labels._

@@

_(B) Comparison of frequencies of traumatic events, manifest PTSD and PTSD symptoms in the study cohort and the cohort of Swiss mountain rescuers reported by Mikutta and colleagues. PTSD symptoms were assessed by the PCL-5 tool and are subsumed under the B, C, B, and E domains. Manifest PTSD is defined as positivity for all domains of the PCL-5 instrument._

@@

```{r fig-mental-literature, fig.width = figur::convert(suppl_figures$mental_literature, to = 'in')$w, fig.height = figur::convert(suppl_figures$mental_literature, to = 'in')$h, fig.cap = 'Comparison of frequency of anxious and depressive symptoms and of resilience scoring in the study cohort and literature.'}

@@

__Supplementary Figure S\@ref(fig:fig-mental-literature). Comparison of frequency of anxious and depressive symptoms and of resilience scoring in the study cohort and literature.__

@@

_(B) Comparison of frequency and severity of depressive symptoms measured by the PHQ-8 tool in the study cohort with 2019 estimates for the general Austrian population._

_(C) Comparison of resilient comping scores in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta et al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.)_

>>>>

Clinically relevant symptoms of anxiety were defined as GAD-7 $\geq$ 10, 
clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 10 [@Humer2022], 
significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 10 [@Kroenke2002]. 
Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008]. 
For a comparison of frequency and severity of depressive symptoms in the study cohort with microcensus data for the general Austrian population, mild depressive symptoms were defined as 5 - 9, moderate symptoms as 10 - 14, moderately severe symptoms as 15 - 19, and severe symptoms as $\geq$ 20 points of the PHQ-8 score [@Kroenke2002a].

@@

Lifetime estimates of prevalence of traumatic events and PTSD in the general population were obtained from representative survey studies [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017]. 
Frequency of traumatic events, manifest PTSD and PTSD symptoms subsumed under the B, C, D, and E domains of the PCL-5 instrument in Swiss mountain rescuers were extracted from the report by Mikutta et al. [@Mikutta2022]. 
Data on frequency of clinically relevant anxious (GAD-7 $\geq$ 10) and depressive (PHQ-9 $\geq$ 10) symptoms in Austria in 2022 were obtained from a publication by Humer et al. [@Humer2022]. 
Mean scores of resilient comping with their standard deviations (SD) obtained with the RS-13 tool were extracted from reports with samples of the German [@Leppert2008] and Tyrolean [@Chernova2021] populations, and from a cohort of Swiss mountain rescuers [@Mikutta2022].

@@

* r statistic: small: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5

* Cramer's V statistic: small: < 0.3, moderate: 0.3 - 0.5, large: $\geq$ 0.5

* $\eta^2$ statistic: small: < 0.06, moderate: 0.06 - 0.14, large: $\geq$ 0.14

* Cohen's $\kappa$ inter-rater reliability statistic: none: < 0.2, minimal: 0.2 - 0.4, small: 0.4 - 0.6, moderate: 0.6 - 0.8, large: $\geq$ 0.8

@@

Anonymized patient data will be made available upon request to the corresponding author. 
The study analysis pipeline is available at https://github.com/PiotrTymoszuk/mental_accident.

@@

flextable::flextable(suppl_tables$aut_population) %>% 
  set_widths(c(3.5, rep(5, 2), 2.8, 2.8)) %>% 
  footnote(7, 1, 
           value = as_paragraph('In the study survey: participants reporting a chronic pre-existing physical illness or a mental disorder diagnosed by a medical professional.'), 
           ref_symbols = 'a', 
           part = 'body') %>%
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic, categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           ref_symbols = 'b', 
           part = 'header') %>% 
  theme_vanilla

@@

```{r fig-accident, fig.width = figur::convert(suppl_figures$accident, to = 'in')$w, fig.height = figur::convert(suppl_figures$accident, to = 'in')$h, fig.cap = 'Age, mountain sport type, and accident month in the study cohort and Austrian national registry of mountain accidents.'}

@@

__Supplementary Figure S\@ref(fig:fig-accident). Age, mountain sport type, and accident month in the study cohort and Austrian national registry of mountain accidents.__

_Age class distribution (A), mountain sport types (B), and accident months (C) were compared between the study cohort and victims of mountain sport accidents in Austria in 2023. Frequencies for the nation-wide data set of mountain accidents were extracted from the annual report of the Austrian Board of Mountain Safety. In the comparison of the age class distribution, only individuals older than 20 years were included. In the comparison of sport types, non-sportive mountain accidents (e.g. during hunting or forestry) were excluded from the national statistic. Non-classifiable sport types (category 'other') in the study cohort were excluded from the analysis presented in (C). Statistical significance of the global differences was determined by $\chi^2$ test with Cramer's V effect size statistic. Pairwise differences for particular sport types, months and age classes were assessed by $\chi^2$ test corrected for multiple testing with the Holm method. Percentages of categories within the complete observation sets are presented as bar plots. Bars are labeled with the corresponding percent values. Total observation numbers are displayed in the plot legends. Global effect sizes and p values are presented in the plot captions. Significant results of post-hoc tests are highlighted with bold X axis labels._

@@

_(B) Comparison of manifest PTSD and PTSD symptoms in the study cohort and the cohort of Swiss mountain rescuers reported by Mikutta and colleagues. PTSD symptoms were assessed by the PCL-5 tool and are subsumed under the B, C, B, and E domains. Manifest PTSD is defined as positivity for all domains of the PCL-5 instrument._

@@

```{r fig-mental-literature, fig.width = figur::convert(suppl_figures$mental_literature, to = 'in')$w, fig.height = figur::convert(suppl_figures$mental_literature, to = 'in')$h, fig.cap = 'Comparison of frequency of clinically relevant symptoms of anxiety and depression, and of resilience scoring in the study cohort and literature.'}

@@

__Supplementary Figure S\@ref(fig:fig-mental-literature). Comparison of frequency of clinically relevant symptoms of anxiety and depression, and of resilience scoring in the study cohort and literature.__

@@

_(C) Comparison of scores of resilient coping in the study cohort, the cohort of Swiss mountain rescuers published by Mikutta et al., and samples of the Tyrolean (Chernova et al.) and German population (Leppert et al.) investigated with the RS13 instrument._
